Shallow EDSLs and Object-Oriented Programming

### 2016-06-14 paper/sections/Overview.tex
Add implicit staging in related work

The new constructs are: \emph{stretch ns c} which inserts additional wires into the circuit \emph{c} by
summing up \emph{ns} and $above\ c_1\ c_2$ which combines two circuits of the same width vertically.
Their types are given below:
\lstinputlisting[language=haskell,linerange=40-41]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES

@@

\lstinputlisting[language=haskell,linerange=40-41]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES
\emph{stretch ns c} inserts additional wires into the circuit \emph{c} by
summing up \emph{ns} and $above c_1 c_2$ combines two circuits of the same width vertically.
With these new constructs, more complex circuits can be constructed.
For example, Figure~\ref{sec:oo} shows the circuit constructed by \emph{stretch [2,2,2] (fan 3) `beside` fan 1}.

To accomplish this task in Haskell, we can simply define two more cases,
\texttt{stretch} and \texttt{above}, for the semantic function:
\lstinputlisting[language=haskell,linerange=45-46]{./code/shallowCircuit.hs}%APPLY:SYNTAX_HS
Similarly, defining two new interfaces
\texttt{Stretch} and \texttt{Above} that both extend \texttt{Circuit} and
implement the \texttt{width} method is all we need to do in Java:

>>>>

The new constructs are: \emph{stretch ns c} which inserts additional wires into the circuit \emph{c} by
summing up \emph{ns} and $above\ c_1\ c_2$ which combines two circuits of the same width vertically.
Their types are given below:
\lstinputlisting[language=haskell,linerange=42-43]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES

@@

\lstinputlisting[language=haskell,linerange=42-43]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES
\emph{stretch ns c} inserts additional wires into the circuit \emph{c} by
summing up \emph{ns} and $above c_1 c_2$ combines two circuits of the same width vertically.
With these new constructs, more complex circuits can be constructed.
For example, Figure~\ref{sec:oo} shows the circuit constructed by \emph{stretch [2,2,2] (fan 3) `beside` fan 1}.

To accomplish this task in Haskell, we can simply define two more cases,
\texttt{stretch} and \texttt{above}, for the semantic function:
\lstinputlisting[language=haskell,linerange=47-48]{./code/shallowCircuit.hs}%APPLY:SYNTAX_HS
Similarly, defining two new interfaces
\texttt{Stretch} and \texttt{Above} that both extend \texttt{Circuit} and
implement the \texttt{width} method is all we need to do in Java:

### 2016-06-14 paper/sections/RelatedWork.tex
Add implicit staging in related work

To enjoy the benefits of both techniques, \cite{svenningsson2012combining}
combines deep and shallow embedding together, which uses shallow embedding to
encode the surface language and translates it to a deeply embedded core language.
Similarly, Yin-Yang~\cite{Jovanovic:2014:YCD:2658761.2658771} allows user
program written in shallow embedding and automatically generates the deeply
embedded version through Scala macros. Hofer and Ostermann~\cite{hofer2010modular} propose to provide both embedding through implementing internal and external visitor at
the same time so that clients can choose for a
particular interpretation.
Our approach, however, can be viewed as a hybrid of shallow and deep embedding -
the encoding looks like shallow embedding while preserving the AST.

@@

In OOP, the Vistor pattern is widely used for modeling languages, which is
well-known for adding new operations easily. Many works
have been done for making it possible to extend variants~\cite{oliveira07genericity,oliveira09modular}.
%In fact, the Visitor pattern has a strong connection with deep and shallow embedding.
% Specifically, the internal variant corresponds to shallow embedding whereas the external variant is equivalent to deep embedding.
Object Algebras~\cite{bruno12oa} simplifies the solution, which is an variant of
internal visitor pattern.
 % However, this approach requires heavy encoding and advanced type system features in Scala.
Another line of work is on OO decomposition. ~\cite{zenger} capture recursive
arguments using virtual types with bounds using Scala. Wang and Oliveira~\cite{eptrivially16}
uses covariant types, a common feature available in most OO languages, to
achieve same goal, which is the basis of our approach.
Compared to the Visitor pattern, the OO decomposition approach is simpler and
our encoding is based on ~\cite{eptrivially16}. The
naming problem for extensions is solved by family polymorphism.

In FP, the two main solutions are \emph{finally
  tagless}~\cite{carette2009finally} and \emph{data types a la
  carte}~\cite{swierstra2008data} (DTC).
Finally tagless approach uses a type class to abstract over all interpretations
of the language. Concrete interpretations are given through creating a data type and
making it an instance of that type.
DTC represents language constructs separately and composes them together using
extensible sums. However, not like OO languages which come with subtyping, one
has to manually implement the subtyping machinery for variants.
Moreover, both DTC and finally tagless approach force dependent interpretations
to be defined along with what they depend on.
This prevents new interpretation that depend on existing interpretations from
being defined modularly while our approach does not have such requirement.

>>>>

To enjoy the benefits of both techniques, Svenningsson and Axelsson~\cite{svenningsson2012combining}
combine deep and shallow embedding together, which uses shallow embedding to
encode the surface language and translates it to a deeply embedded core language.
Similarly, \emph{Yin-Yang}~\cite{Jovanovic:2014:YCD:2658761.2658771} allows user
program written in shallow embedding and automatically generates the deeply
embedded version through Scala macros. Hofer and Ostermann~\cite{hofer2010modular} propose to provide both embedding through implementing internal and external visitor at
the same time so that clients can choose for a particular interpretation.
Scherr and Chiba~\cite{scherr2014implicit} uses \emph{implicit staging} to
extract shallow EDSL part from a normal program and represent it in a form
similar to AST, allowing optimizations to be performed.
Our approach, however, supports transformations in the first place while retaining the
simplicity of the shallow embedding.

@@

In OOP, the Vistor pattern is widely used for modeling languages, which is
well-known for adding new operations easily. Many works
have been done for making it possible to extend variants~\cite{oliveira07genericity,oliveira09modular}.
%In fact, the Visitor pattern has a strong connection with deep and shallow embedding.
% Specifically, the internal variant corresponds to shallow embedding whereas the external variant is equivalent to deep embedding.
Object Algebras~\cite{bruno12oa} simplifies the solution, which is an variant of
internal visitor pattern.
 % However, this approach requires heavy encoding and advanced type system features in Scala.
Another line of work is on OO decomposition. ~\cite{zenger05independentlyextensible} capture recursive
arguments using virtual types with bounds using Scala. Wang and Oliveira~\cite{eptrivially16}
use covariant type refinements, a common feature available in conventional OO
languages, to solve the EP, which is the basis of our approach.
Compared to the Visitor pattern, the OO decomposition approach is simpler and
our encoding is based on ~\cite{eptrivially16}. The
naming problem for extensions is solved by family polymorphism.

In FP, the two main solutions are \emph{finally
  tagless}~\cite{carette2009finally} and \emph{data types a la
  carte}~\cite{swierstra2008data} (DTC).
Finally tagless approach uses a type class to abstract over all interpretations
of the language. Concrete interpretations are given through creating a data type and
making it an instance of that type.
DTC represents language constructs separately and composes them together using
extensible sums. However, not like OO languages which come with subtyping, one
has to manually implement the subtyping machinery for variants.
Moreover, both DTC and finally tagless approach force dependent interpretations
to be defined along with what they depend on.
This prevents new interpretation that depend on existing interpretations from
being defined modularly.

### 2016-06-14 paper/sections/ShallowOO.tex
Add implicit staging in related work

In this section, we argue that \emph{shallow embedding} and straightforward OO
programming are closely related. To demonstrate it, we use a subset
of the DSL presented in Gibbons and Wu's paper~\cite{} as the running example.
We first give the original implementation of the language in shallow embedding
and then gradually rewrite the definition into an ``OO style''. Also, we show how to do
transformations in shallow embedding, which is typically thought to be the
privilege of deep embedding.

@@

There are three constructs - two primitives
\emph{identity} and \emph{fan} and one combinator \emph{beside}, whose
meanings are: \emph{identity n} contains \emph{n} isolated vertical wires;
\emph{fan n} contains \emph{n} vertical wires and the first wire connects to
all the remaining wires from top to bottom; $beside\ c_1\ c_2$ joins two circuits
$c_1$ and $c_2$ horizontally.
For example, the circuit \emph{fan 3 beside identity 3} can be visualized in Figure~\ref{}.

@@

\lstinputlisting[language=haskell,linerange=2-2]{./code/shallowCircuit.hs}%APPLY:CIRCUIT_TYPE
\lstinputlisting[language=haskell,linerange=5-7]{./code/shallowCircuit.hs}%APPLY:TYPES
The type \texttt{Circuit}, representing the semantic domain, is not defined yet
and should be filled in later with a concrete type according to the semantics.
For example, if the semantics of \dsl is to calculate the width of a
circuit, the semantic function would be:
\lstinputlisting[language=haskell,linerange=11-16]{./code/shallowCircuit.hs}%APPLY:CIRCUIT1
As shallow embedding is semantics-oriented, it is already an integer after
constructing a circuit and hence \texttt{width1} is an identity function.
For example, we will immediately get $6$ for
\texttt{fan 3 `beside` identity 3} and hence \texttt{width1} is just an identity function.

The above implementation can be rewritten by wrapping the result into an
datatype and get back the value through pattern matching:
\lstinputlisting[language=haskell,linerange=20-26]{./code/shallowCircuit.hs}%APPLY:CIRCUIT2
Then circuit construction gives back a \texttt{Circuit2} which contains an
integer value representing the width and \texttt{width2} extracts out the value.
%One may be curious whether this variant implementation is still shallow embedding or not. The answer is yes because ... % why it is shallow

The rewrite goes on by using a record with a sole field to represent the result:
\lstinputlisting[language=haskell,linerange=31-36]{./code/shallowCircuit.hs}%APPLY:CIRCUIT3
We no longer need a separate \texttt{width3} function as the field name is the extractor.
The implementation is still shallow since essentially it is nothing different
from the last one.
However, the definition already approaches an OO program such that we can easily
translate it into an OO language like Java:
\lstinputlisting[linerange=7-26]{../src/paper/sec2/Circuit.java}%APPLY:CIRCUIT_OO
where the record type maps to the abstract class \texttt{Circuit} and field
declaration becomes method declaration.
Each case in the semantic function corresponds to an class and its parameters become fields of the class.
And all the classes extends \texttt{Circuit} by giving \texttt{width} a definition.

This implementation is exactly how we model \dsl in an OO language for the first
place, following the \interp pattern.
In other words, shallow embedding and straightforward OO programming are closely
related (deep embedding is closely related to the \textsc{Visitor} pattern).

@@

An instance of this kind of operations could be desugaring, which eliminates some
language constructs that could be expressed by other language constructs without
sacrificing the expressiveness of the language.
In \dsl, the \emph{identity n} construct is such a syntactic sugar that can
be rewritten via the following formula:
$$identity\ n = (fan\ 1) \overbrace{\ beside\ (fan\ 1)}^{\text{repeat }n-1\text{ times}}\ (n > 0)$$
which states that \emph{identity n} can be represented as \emph{n} of \emph{fan 1} combined with \emph{beside}.
%Using the above formula, we can define a transformation that desugars every
%occurrence of \emph{identity} to a combination of \emph{fan} and \emph{beside}.

>>>>

In this section, we argue that \emph{shallow embedding} and straightforward OO
programming are closely related. To demonstrate it, we use a subset
of the DSL presented in Gibbons and Wu's paper~\cite{gibbons2014folding} as the running example.
We first give the original shallow embedded implementation of the language
written in Haskell
and gradually rewrite it approaching to an ``OO style''.
Then translating the program into an OO version like Java becomes straightforward.
We also illustrate how to do transformations in shallow embedding, which is typically regarded as the privilege of deep embedding.

@@

There are three constructs - two primitives
\emph{identity} and \emph{fan} and one combinator \emph{beside}, whose
meanings are: \emph{identity n} contains \emph{n} isolated vertical wires;
\emph{fan n} contains \emph{n} vertical wires and the first wire connects to
all the remaining wires from top to bottom; $beside\ c_1\ c_2$ joins two circuits
$c_1$ and $c_2$ horizontally.
For example, the circuit \emph{beside (fan 3) (identity 3)} can be visualized in Figure~\ref{}.

@@

\lstinputlisting[language=haskell,linerange=2-2]{./code/shallowCircuit.hs}%APPLY:CIRCUIT_TYPE
\lstinputlisting[language=haskell,linerange=5-7]{./code/shallowCircuit.hs}%APPLY:TYPES
The type \texttt{Circuit}, representing the semantic domain, is not defined yet
and should be filled in later with a concrete type according to the semantics.
For example, if the semantics of \dsl is to calculate the width of a
circuit, the semantic function would be:
\lstinputlisting[language=haskell,linerange=11-17]{./code/shallowCircuit.hs}%APPLY:CIRCUIT1
As shallow embedding is semantics-oriented, it is already an integer after
constructing a circuit and hence \texttt{width1} is an identity function.
For example, we will immediately get $6$ for
\texttt{beside (fan 3) (identity 3)} and hence \texttt{width1} is just an identity function.

The above implementation can be rewritten by wrapping the result into an
datatype and get back the value through pattern matching:
\lstinputlisting[language=haskell,linerange=21-28]{./code/shallowCircuit.hs}%APPLY:CIRCUIT2
% Then circuit construction gives back a \texttt{Circuit2} which contains an
% integer value representing the width and \texttt{width2} extracts out the value.
% Is this definition still shallow embedding?
%One may be curious whether this variant implementation is still shallow embedding or not. The answer is yes because ... % why it is shallow

The rewrite goes on by using a record with a sole field to represent the result:
\lstinputlisting[language=haskell,linerange=33-38]{./code/shallowCircuit.hs}%APPLY:CIRCUIT3
We no longer need a separate \texttt{width3} function as the field name is the extractor.
The implementation is still shallow since essentially it is nothing different
from the last one.
However, the definition already approaches an OO program such that we can easily
translate it into an OO language like Java:
\lstinputlisting[linerange=7-26]{../src/paper/sec2/Circuit.java}%APPLY:CIRCUIT_OO
where the record type maps to the abstract class \texttt{Circuit} and field
declaration becomes method declaration.
Each case in the semantic function corresponds to an class and its parameters become fields of the class.
And all the classes extends \texttt{Circuit} by giving \texttt{width} a definition.

This implementation is exactly how we would model \dsl in an OO language in the first
place, following the \interp pattern.
In other words, shallow embedding and straightforward OO programming are closely
related.
It may worth mentioning that deep embedding is closely related to the \textsc{Visitor} pattern.

@@

An instance of this kind of operations could be desugaring, which eliminates some
language constructs that could be expressed by other language constructs without
sacrificing the expressiveness of the language.
In \dsl, the \emph{identity n} construct is such a syntactic sugar that can
be rewritten via the following formula:
$$identity\ n = \overbrace{\ beside\ (fan\ 1)}^{\text{repeat }n-1\text{ times}}\ (fan\ 1)\ (n > 0)$$
which states that \emph{identity n} can be represented as \emph{n} of \emph{fan 1} combined with \emph{beside}.
%Using the above formula, we can define a transformation that desugars every
%occurrence of \emph{identity} to a combination of \emph{fan} and \emph{beside}.
